---
title: ""
message: false
warning: false
---

# Dlnm approach

## Packages, options and parameters

This chunk prepares the computational environment and defines the global structural parameters that govern the Distributed Lag Non‑Linear Model (DLNM) and the subsequent inferential model. It does not yet fit a statistical model; instead, it establishes the methodological framework under which the DLNM is constructed and later estimated within a generalized (mixed) model.

```{r}


  library(dplyr)
  library(tidyr)
  library(readr)
  library(readxl)
  library(dlnm)
  library(brms)
  library(ggplot2)
  library(posterior)
  library(purrr)


set.seed(20251218)

# ------------------------------- PARÂMETROS ----------------------------------
END_DAP   <- 85     # 80, 85 ou 90 (ponto de avaliação, p.ex. ~R6)
LAG_MAX   <- 85     # lags máximos (dias antes de END_DAP)
DF_VAR    <- 4      # flexibilidade na dimensão de exposição (ns)
DF_LAG    <- 4      # flexibilidade na dimensão de lag (ns)
SEPARATOR <- LAG_MAX

VARS <- c("tmax","rain_cum","vpd")

LAG_BAND_A <- 0:40
LAG_BAND_B <- 41:LAG_MAX
```

## Data importation and pre organization

To ensure valid inference using a Distributed Lag Non‑Linear Model (DLNM), epidemiological observations and weather time series were first aligned in time and space and filtered to guarantee complete exposure histories. Disease severity was evaluated at a fixed assessment time (END_DAP), corresponding approximately to the R6 growth stage, and each epidemic was linked to its assumed assessment date. Daily weather data were then restricted to the period from planting (day 0) to the assessment day, ensuring that all lagged exposures used in the DLNM were fully observed. Only epidemics with complete daily weather records across the entire lag window were retained, as DLNM estimation requires a continuous exposure history for all lags considered. Vapor pressure deficit (VPD) was derived from mean temperature and relative humidity to represent atmospheric drying demand and leaf wetness persistence, which are epidemiologically more meaningful drivers of foliar disease development than relative humidity alone. A site–year identifier was constructed to capture unobserved spatial and temporal heterogeneity in subsequent mixed‑effects modeling. Finally, because disease severity was modeled using a beta regression, observations equal to 0 or 1 were slightly adjusted to satisfy the distributional assumptions of the beta likelihood. This preprocessing step establishes a consistent epidemiological time origin, guarantees complete lagged exposure information, and produces physically interpretable climatic covariates, thereby providing a sound foundation for constructing DLNM cross‑basis functions and estimating their effects within a generalized mixed modeling framework.

## ERA5 weather extraction
```{r}

epi <- readxl::read_xlsx("data/ts_check.xlsx")

epi = epi %>% 
  mutate(
    PD85 = planting_date+85
  )

epi$planting_date = as.Date(epi$planting_date)

epi = epi %>% 
  mutate(
    PD85 = planting_date + 85
  )

epi_use <- epi |>
  rename(
    sev = mean_sev
  ) |>
  filter(!is.na(sev)) |>
  mutate(
    planting_date = as.Date(planting_date, origin = "1899-12-30"),
    epi_id = row_number(),
    siteyear = paste0(location, "_", year)
  )

epi_use
```

```{r, echo=TRUE, eval=FALSE}
wx = get_era5(
  data = epi_use,
  days_around = "85",
  date_col = "planting_date",
  study_col = "epi_id",
  pars = c("temperature_2m", "relative_humidity_2m", "precipitation", "dewpoint_2m"),
  models = NULL,
  direction = "forth"
)


colnames(wx) = c("planting_date", "tmean","tmax","tmin","rh","rain","t2mdew","longitude","latitude","epi_id")


unique(wx$epi_id)

wx_use <- epi_use |>
  select(study,year,location,epi_id) %>% 
  left_join(wx, by = "epi_id")

wx_use$dpp <- rep(0:85, times = 254)


es <- function(T) { 0.6108 * exp((17.27 * T) / (T + 237.3)) }
wx_use <- wx_use |>
  mutate(vpd = es(tmean) * (1 - rh / 100))

#wx_use = wx_use %>% 
  #filter(dpp >=15)


wx_use <- wx_use |>
  arrange(epi_id, planting_date) |>
  group_by(epi_id) |>
  mutate(rain_cum = cumsum(rain)) |>
  ungroup()

#writexl::write_xlsx(wx_use,"data/ERA5_weather.xlsx")
```

```{r}
wx_use = read_xlsx("data/ERA5_weather.xlsx")
```

## Templates to cross-basis (shared)

To model delayed and non-linear effects of climatic variables on disease severity, we adopted the Distributed Lag Non-Linear Model (DLNM) framework. Because disease severity is observed only once per epidemic—at a fixed assessment time—while climatic exposures are recorded daily, the DLNM requires a complete and consistently structured exposure history for each epidemic across the entire lag window.

For each climatic variable, daily exposure series were pooled across epidemics and concatenated with artificial separation gaps. These gaps prevent the propagation of lag effects from one epidemic into another, ensuring that temporal dependence is preserved within epidemics but not across them. This pooled representation allows the construction of a single shared cross-basis template per variable, guaranteeing that the same exposure–lag functional form is used for all epidemics.

Cross-basis templates were defined using natural cubic splines in both the exposure and lag dimensions. The degrees of freedom controlling exposure flexibility (`DF_VAR`) and lag flexibility (`DF_LAG`) determine the smoothness and complexity of the response surface but do not estimate any effects at this stage. Instead, these templates define the functional space within which the distributed lag effects are later estimated inside a generalized mixed modeling framework.

By constructing shared cross-basis templates prior to model fitting, this step ensures structural consistency, avoids repeated redefinition of spline geometries, and facilitates epidemiologically interpretable comparisons of climatic effects across lag periods and variables.

Let $i = 1, \dots, N$ index epidemics, and let $t_i$ denote the disease assessment time for epidemic $i$. For a climatic variable $v$ (e.g., rainfall, mean temperature, or vapor pressure deficit), let $x_{iv}(t_i - l)$ represent the exposure at lag $l$, where $l = 0, 1, \dots, L$ and $L$ is the maximum lag considered.

The DLNM assumes that the contribution of variable $v$ to the linear predictor can be written as a smooth bivariate function of exposure and lag:

$$
f_v(x, l)
=
\sum_{k=1}^{K_v}
\sum_{m=1}^{M_v}
\beta^{(v)}_{km}
\, B^{(v)}_k(x)
\, C^{(v)}_m(l),
$$

where:

-   $B^{(v)}_k(x)$ are spline basis functions for the exposure dimension, with $K_v = \text{DF\_VAR}$,
-   $C^{(v)}_m(l)$ are spline basis functions for the lag dimension, with $M_v = \text{DF\_LAG}$,
-   $\beta^{(v)}_{km}$ are coefficients to be estimated in the inferential model.

The total contribution of variable $v$ for epidemic $i$ is obtained by summing across all lags:

$$
\eta_i^{(v)}
=
\sum_{l=0}^{L}
f_v\!\left(x_{iv}(t_i - l),\, l\right).
$$

These contributions enter the full model additively. For example, using a logit link for beta regression, the mean disease severity $\mu_i$ is modeled as:

$$
\text{logit}(\mu_i)
=
\beta_0
+
\sum_v \eta_i^{(v)}
+
u_{\text{siteyear}(i)},
$$

where $u_{\text{siteyear}(i)} \sim \mathcal{N}(0, \sigma^2)$ represents a random intercept accounting for site–year–level heterogeneity.

```{r}

# ---- Chunk 2: cross-basis templates ----------------------------------------
build_pooled_series <- function(wx_long, var, sep_n = SEPARATOR) {
  ids <- unique(wx_long$epi_id)
  out <- vector("list", length(ids))
  for (i in seq_along(ids)) {
    v <- wx_long |> filter(epi_id == ids[i]) |> arrange(dpp) |> pull(.data[[var]])
    out[[i]] <- c(v, rep(NA_real_, sep_n))
  }
  unlist(out)
}

cb_templates <- list()
for (v in VARS) {
  stopifnot(v %in% names(wx_use))
  x_pool <- build_pooled_series(wx_use, v, sep_n = SEPARATOR)
  cb_templates[[v]] <- dlnm::crossbasis(
    x_pool, lag = LAG_MAX,
    argvar = list(fun = "ns", df = DF_VAR),
    arglag = list(fun = "ns", df = DF_LAG)
  )
}

```

## Epidemic matrix (design)

Disease severity was observed once per epidemic at the assessment time, whereas climatic exposures were recorded daily over the entire growing period. To reconcile these different temporal resolutions within the DLNM framework, the lagged exposure histories of each epidemic were mapped onto epidemic-level covariates using pre-defined cross-basis templates.

For each climatic variable, the daily exposure series of an epidemic was transformed into a vector of basis coefficients corresponding to the last observed time point (the assessment day). This operation collapses the full distributed lag structure into a single row per epidemic while preserving the cumulative contribution of past exposures encoded by the DLNM. By applying the same cross-basis template to all epidemics, this step ensures that each epidemic is represented within a common functional space, allowing the resulting coefficients to be estimated consistently within a generalized mixed-effects model.

The resulting epidemic-level design matrix contains one row per epidemic and multiple columns per climatic variable, each column corresponding to a specific exposure–lag basis function. These columns serve as fixed-effect predictors in the inferential model, linking past climatic conditions to disease severity through the DLNM representation.

Let $i = 1, \dots, N$ index epidemics, and let $t_i$ denote the disease assessment time for epidemic $i$. For a climatic variable $v$, let $x_{iv}(t_i - l)$ be the exposure at lag $l$, with $l = 0, 1, \dots, L$.

Using the previously defined cross-basis functions, the distributed lag contribution of variable $v$ for epidemic $i$ can be written as:

$$
\eta_i^{(v)}
=
\sum_{l=0}^{L}
f_v\!\left(x_{iv}(t_i - l),\, l\right),
$$

where the exposure–lag response surface is approximated by a tensor product of spline bases:

$$
f_v(x,l)
=
\sum_{k=1}^{K_v}
\sum_{m=1}^{M_v}
\beta^{(v)}_{km}
\, B^{(v)}_k(x)
\, C^{(v)}_m(l).
$$

Here, $B^{(v)}_k(x)$ are the spline basis functions for the exposure dimension and $C^{(v)}_m(l)$ are the spline basis functions for the lag dimension. The coefficients $\beta^{(v)}_{km}$ are estimated in the inferential model.

Evaluating the cross-basis at the final time point $t_i$ yields a vector of basis values for each epidemic, denoted $\mathbf{X}_i^{(v)}$. Stacking these vectors across all climatic variables produces the epidemic-level design matrix:

$$
\mathbf{X}_i
=
\left[
\mathbf{X}_i^{(\text{rain_cum})},
\mathbf{X}_i^{(\text{tmax})},
\mathbf{X}_i^{(\text{vpd})}
\right].
$$

This design matrix enters the generalized mixed-effects model as fixed effects:

$$
\text{logit}(\mu_i)
=
\beta_0
+
\mathbf{X}_i \boldsymbol{\beta}
+
u_{\text{siteyear}(i)},
$$

where $\mu_i$ is the mean disease severity for epidemic $i$ and $u_{\text{siteyear}(i)} \sim \mathcal{N}(0,\sigma^2)$ accounts for site–year–level heterogeneity.

```{r}

# ---- Chunk 3: per-epidemic design (one row per epidemic) -------------------
extract_last_cb_row <- function(x, cb_template) {
  cb <- dlnm::crossbasis(
    x, lag = LAG_MAX,
    argvar = attr(cb_template, "argvar"),
    arglag = attr(cb_template, "arglag")
  )
  as.numeric(cb[length(x), ])
}

build_design_for_var <- function(wx_long, var, cb_template, prefix) {
  X <- wx_long |>
    group_by(epi_id) |>
    summarise(cb = list(extract_last_cb_row(.data[[var]], cb_template)),
              .groups = "drop")
  p  <- length(X$cb[[1]])
  nm <- paste0(prefix, seq_len(p))
  X  |> mutate(cb = lapply(cb, setNames, nm)) |> tidyr::unnest_wider(cb)
}

X_list <- map(VARS, ~ build_design_for_var(wx_use, .x, cb_templates[[.x]],
                                           prefix = paste0("cb_", .x, "_")))
names(X_list) <- VARS

X <- reduce(X_list, left_join, by = "epi_id")

dat <- epi_use |>
dplyr::select(epi_id, sev, siteyear, year, location, state, latitude, longitude, planting_date, PD85) |>
  left_join(X, by = "epi_id")


#dat <- epi_use |>
#dplyr::select(epi_id, sev, siteyear, year, location, state, lat, longitude, planting_date) |>
#  left_join(X, by = "epi_id")
```

This step converts the full daily exposure histories into epidemic-level predictors that retain the cumulative and delayed effects of climate encoded by the DLNM. The transformation allows the distributed lag structure to be estimated using a single observation per epidemic while maintaining biological interpretability of the climatic effects across time.

## Framework modeling

### Inferential framework and model fitting

After constructing epidemic-level DLNM design matrices, disease severity was modeled using a generalized mixed-effects framework. Because the response variable represents a proportion bounded between 0 and 1, a beta regression with a logit link was adopted. This formulation allows flexible modeling of the mean disease severity while respecting the bounded nature of the data.

The DLNM cross-basis coefficients derived for each climatic variable were included as fixed effects in the linear predictor, linking past climatic exposures to disease severity through their distributed and non-linear effects. To account for unobserved heterogeneity among epidemics conducted at different locations and years, a random intercept was specified at the site–year level. This hierarchical structure captures baseline differences in disease pressure that are not explained by the climatic covariates.

Importantly, the DLNM itself does not perform inference; rather, it defines the structure of the predictors. All statistical inference—including estimation of coefficients, uncertainty quantification, and hypothesis testing—is carried out within the mixed-effects beta regression model. Thus, this step represents the point at which the distributed lag non-linear effects are formally estimated and interpreted.

Let $i = 1, \dots, N$ index epidemics, and let $y_i \in (0,1)$ denote the observed disease severity at the assessment time. Disease severity is assumed to follow a beta distribution with mean $\mu_i$ and precision parameter $\phi$:

$$
y_i \sim \text{Beta}(\mu_i, \phi).
$$

The mean $\mu_i$ is linked to a linear predictor via a logit link function:

$$
\text{logit}(\mu_i)
=
\eta_i.
$$

The linear predictor is specified as:

$$
\eta_i
=
\beta_0
+
\sum_{v}
\mathbf{X}_i^{(v)} \boldsymbol{\beta}^{(v)}
+
u_{\text{siteyear}(i)},
$$

where:

-   $\beta_0$ is the global intercept,\
-   $\mathbf{X}_i^{(v)}$ is the epidemic-level DLNM design vector for climatic variable $v$,\
-   $\boldsymbol{\beta}^{(v)}$ is the corresponding vector of fixed-effect coefficients,\
-   $u_{\text{siteyear}(i)} \sim \mathcal{N}(0, \sigma^2)$ is a random intercept capturing site–year–specific deviations.

Each design vector $\mathbf{X}_i^{(v)}$ encodes the cumulative contribution of past exposures through the DLNM cross-basis evaluated at the assessment time:

$$
\mathbf{X}_i^{(v)} \boldsymbol{\beta}^{(v)}
=
\sum_{l=0}^{L}
f_v\!\left(x_{iv}(t_i - l),\, l\right),
$$

where $f_v(\cdot,\cdot)$ denotes the exposure–lag response surface defined by the DLNM for variable $v$.

```{r}


# ---- Chunk 4: fit brms beta hierarchical model ------------------------------
cb_terms <- paste(names(dat)[grepl("^cb_", names(dat))], collapse = " + ")
fml <- as.formula(paste0("sev ~ 1 + (1|siteyear) + ", cb_terms))

library(glmmTMB)
dat$sev  = dat$sev/100
  
  
fit_tmb <- glmmTMB( #fit_tmb
  formula = fml,
  data = dat,
  family = beta_family(link = "logit")
)


summary(fit_tmb)
```

This modeling step represents the inferential core of the analysis. The DLNM defines how climatic effects are structured across exposure levels and time lags, while the mixed-effects beta regression estimates the magnitude, direction, and uncertainty of these effects. The resulting model quantifies how past climatic conditions jointly shape disease severity at the assessment time, accounting for both delayed responses and unobserved heterogeneity among site–year combinations.

### Model performance

After fitting the DLNM-based mixed-effects beta regression model, fitted values were obtained for each epidemic to quantify the model-implied mean disease severity at the assessment time. These fitted values represent the expected severity conditional on the estimated distributed lag non-linear effects of climatic variables and the random site–year effects.

Predictions were generated on the response scale, yielding values directly interpretable as mean disease severity proportions. By attaching these fitted values to the epidemic-level dataset, this step facilitates subsequent model evaluation, diagnostic checks, and graphical comparison between observed and model-implied disease severity. Importantly, these predictions are not used to refit the model; rather, they summarize how the inferred DLNM structure translates into expected epidemiological outcomes for the observed data.

Let $y_i \in (0,1)$ denote the observed disease severity for epidemic $i$, and let $\mu_i$ be its model-implied mean. After fitting the mixed-effects beta regression model, predictions are obtained as:

$$
\hat{\mu}_i
=
\mathbb{E}(y_i \mid \mathbf{X}_i, u_{\text{siteyear}(i)}),
$$

where $\mathbf{X}_i$ is the epidemic-level DLNM design vector and $u_{\text{siteyear}(i)}$ is the estimated random intercept for the corresponding site–year.

Using a logit link, the predicted mean is given by:

$$
\hat{\mu}_i
=
\text{logit}^{-1}(\hat{\eta}_i),
$$

with linear predictor:

$$
\hat{\eta}_i
=
\hat{\beta}_0
+
\sum_v \mathbf{X}_i^{(v)} \hat{\boldsymbol{\beta}}^{(v)}
+
\hat{u}_{\text{siteyear}(i)}.
$$

Here, $\hat{\beta}_0$ and $\hat{\boldsymbol{\beta}}^{(v)}$ are the estimated fixed-effect coefficients, and $\hat{u}_{\text{siteyear}(i)}$ denotes the best linear unbiased predictor (BLUP) of the random effect.

The resulting fitted values $\hat{\mu}_i$ correspond to the expected disease severity at the assessment time, given the full distributed lag history of climatic exposures encoded by the DLNM.

```{r}

t = predict(fit_tmb, newdata = dat, type = "response")


dat2 = dat
dat2$t = t


```

This step translates the inferred DLNM effects into epidemic-level expectations of disease severity on the natural response scale. The fitted values summarize how the combined, lagged influence of climate and site–year heterogeneity manifests in expected disease outcomes, providing a direct link between the statistical model and epidemiological interpretation.

#### Agreement between observed and model-implied severity (CCC decomposition)

To quantify how well the fitted model reproduces observed disease severity on the response scale, we evaluated agreement between observed severities ($y_i$) and fitted mean severities ($\hat{\mu}_i$) using Lin’s Concordance Correlation Coefficient (CCC). CCC measures both (i) *precision* (how tightly predictions follow the observed variability) and (ii) *accuracy*/*bias* (how close predictions are to the 1:1 line). This provides a single interpretable agreement metric that complements visual diagnostics.

We also report the two standard CCC components: the Pearson correlation (precision component) and the bias correction factor (accuracy component). Together, these summarize whether lack of agreement is driven primarily by random dispersion (low precision), systematic bias (low accuracy), or both.

Let $y_i$ denote observed severity and $\hat{\mu}_i$ denote the model-implied mean severity for epidemic $i$. Define:

-   sample means: $\bar{y}$ and $\overline{\hat{\mu}}$
-   sample variances: $s_y^2$ and $s_{\hat{\mu}}^2$
-   sample covariance: $s_{y,\hat{\mu}}$

The Pearson correlation (precision component) is:

$$
\rho
=
\frac{s_{y,\hat{\mu}}}{s_y\, s_{\hat{\mu}}}.
$$

Lin’s concordance correlation coefficient is:

$$
\rho_c
=
\frac{2 s_{y,\hat{\mu}}}{s_y^2 + s_{\hat{\mu}}^2 + (\bar{y}-\overline{\hat{\mu}})^2}.
$$

It can be decomposed as:

$$
\rho_c = \rho \times C_b,
$$

where $C_b$ is the bias correction factor (accuracy component), with $0 < C_b \le 1$, penalizing deviations from the 1:1 line due to differences in location and scale.

```{r}

library(epiR)

tt = epi.ccc(dat2$sev,dat2$t)

tt$rho.c

tt$C.b

r <- tt$rho.c[,1]/tt$C.b
r

gc()
```

#### Graphical calibration and residual error distribution

To complement the CCC summary, we visualized model calibration by plotting observed severity ($y_i$) against fitted mean severity ($\hat{\mu}_i$) with a 1:1 reference line. Points close to this line indicate good calibration; systematic deviations reflect bias (e.g., underprediction at high severities).

We additionally computed residual errors on the response scale and examined their empirical distribution via a histogram, overlaying a normal density with matching mean and standard deviation. While the beta regression model does not assume normal residuals on the response scale, this visualization provides a simple check for gross asymmetry, heavy tails, or systematic shifts (non-zero mean error) that may indicate calibration issues.

For epidemic $i$, define the response-scale residual error as:

$$
e_i = y_i - \hat{\mu}_i.
$$

The mean residual error summarizes average bias:

$$
\bar{e} = \frac{1}{N}\sum_{i=1}^{N} e_i,
$$

and the residual spread is summarized by:

$$
s_e = \sqrt{\frac{1}{N-1}\sum_{i=1}^{N}(e_i - \bar{e})^2 }.
$$

Calibration can be visually assessed by the proximity of points $(y_i, \hat{\mu}_i)$ to the 1:1 line:

$$
\hat{\mu} = y.
$$

```{r}

CCC  <- 0.89
Bias <- 0.94
Precision <- 0.94

library(scales)
lab <- paste0(
  "CCC = ", number(CCC, accuracy = 0.01, decimal.mark = "."),
  "\nBias = ", number(Bias, accuracy = 0.01, decimal.mark = "."),
  "\nPrecision = ", number(Precision, accuracy = 0.01, decimal.mark = ".")
)


model = dat2 %>% 
  ggplot(aes(sev,t))+
  geom_jitter(size = 2)+
  #geom_smooth(method = "lm", se = F, color = "orange")+
   geom_abline(slope = 1, intercept = 0, color = "tan2",
              linetype = "solid", size = 1.4) +
  scale_x_continuous(breaks=c(0.00, 0.25, 0.50,0.75), limits = c(0,0.75))+
  scale_y_continuous(breaks=c(0.00, 0.25, 0.50,0.75), limits = c(0,0.75))+
  theme_bw()+
  coord_cartesian(clip = "off")+
  labs(x = "Observed (%)",
       y = "Predicted (%)")+
  annotate(
    "text",
    x = -Inf, y = Inf,
    label = lab,
    hjust = -0.05,   
    vjust = 1.1,     
    size = 3.5,
    fontface = "bold",
    lineheight = 1.05
  )


dat2 = dat2 %>% 
  mutate(
    error = sev - t
  )

mean_res = mean(dat2$error)
sd_res = sd(dat2$error)



hist = dat2 %>% 
  ggplot(aes(error))+
  geom_histogram(aes(y = ..density..), fill = "black", color = "white", bins = 10) + # Ajusta para densidade
  stat_function(fun = dnorm, args = list(mean = mean_res, sd = sd_res), 
                color = "orange", size = 1.2, linetype = "solid")+
  theme_bw()+
 labs(
    x = "Residual error",
    y = "Frequency"
  )

```

#### Simulation-based residual diagnostics (DHARMa)

Because beta mixed models can exhibit complex residual behavior—particularly under non-Gaussian likelihoods and random effects—we evaluated model adequacy using simulation-based residual diagnostics implemented in DHARMa. DHARMa simulates response replicates from the fitted model (including the specified distribution and random effects structure) and constructs *scaled residuals* that should be approximately Uniform(0,1) under correct model specification.

We then assessed the uniformity of the simulated residuals using a quantile–quantile (QQ) comparison between observed residual quantiles and their theoretical Uniform(0,1) quantiles. Systematic departures from the 1:1 line indicate potential misspecification, such as incorrect distributional assumptions, unmodeled structure, or dispersion issues.

Let $y_i$ be the observed response. DHARMa generates $M$ simulated responses:

$$
y_i^{(1)}, y_i^{(2)}, \dots, y_i^{(M)} \sim p(\cdot \mid \hat{\theta}),
$$

where $\hat{\theta}$ denotes the fitted model parameters (including fixed effects, random effects structure, and distributional parameters). The scaled residual is constructed using the empirical cumulative distribution of the simulations:

$$
u_i
=
\frac{1}{M}\sum_{m=1}^{M} \mathbb{I}\!\left(y_i^{(m)} \le y_i\right),
$$

which should satisfy:

$$
u_i \sim \text{Uniform}(0,1)
\quad \text{if the model is correctly specified.}
$$

For a QQ plot, the theoretical Uniform(0,1) quantiles are:

$$
q
$$

```{r}

library(DHARMa)
library(ggplot2)

res <- simulateResiduals(fit_tmb)


u <- sort(res$scaledResiduals)
n <- length(u)


theoretical <- ((1:n) - 0.5) / n


df <- data.frame(
  theoretical = theoretical,
  observed = u
)


qq_plot = ggplot(df, aes(x = theoretical, y = observed)) +
  geom_jitter(color = "black", alpha = 0.7, size = 3) +
  geom_abline(slope = 1, intercept = 0, color = "orange", linewidth = 1) +
  labs(
    x = "Theoritical quantis",
    y = "Simulated residues"
  )+
  theme_bw()



```

```{r}
model|qq_plot/hist

ggsave("fig/model.png", dpi = 600, width = 8, height = 6)
```

### Estimation of cumulative relative risks over epidemiologically defined lag windows

To summarize the distributed effects of climatic variables in an epidemiologically interpretable manner, relative risks were derived from the fitted DLNM by aggregating lag-specific contributions over predefined temporal windows. Rather than interpreting individual lag effects, which may be difficult to relate directly to biological processes, cumulative effects were computed for early and late lag periods corresponding to distinct phases of epidemic development.

For each climatic variable, the DLNM provides an estimated exposure–lag response surface describing how past exposures contribute to disease severity at the assessment time. By summing the lag-specific contributions across a given lag window and exponentiating the resulting linear predictor, a cumulative relative risk is obtained. This quantity represents the proportional change in expected disease severity associated with a given exposure level, relative to a reference value, accounting for the entire exposure history within the specified lag interval.

Lag windows were defined a priori to reflect biologically meaningful phases of the epidemic. The early window (0–40 days before assessment) captures processes related to infection establishment, latent development, and early lesion formation, whereas the late window (41–70 days) reflects symptom expansion and disease expression near the assessment stage. This aggregation facilitates direct comparison of climatic effects across variables and epidemic phases, while preserving the non-linear and delayed structure estimated by the DLNM.

Let $i$ index epidemics and let $t_i$ denote the disease assessment time. For a climatic variable $v$, let $x_{iv}(t_i - l)$ be the exposure at lag $l$, with $l = 0, 1, \dots, L$.

The DLNM defines an exposure–lag response surface $f_v(x,l)$, such that the total contribution of variable $v$ to the linear predictor is:

$$
\eta_i^{(v)}
=
\sum_{l=0}^{L}
f_v\!\left(x_{iv}(t_i - l),\, l\right).
$$

To obtain a cumulative effect over a specific lag window $\mathcal{L} \subset \{0,\dots,L\}$, the contribution is restricted to that window:

$$
\eta_{i,\mathcal{L}}^{(v)}
=
\sum_{l \in \mathcal{L}}
f_v\!\left(x_{iv}(t_i - l),\, l\right).
$$

Relative risk is then defined with respect to a reference exposure value $x_v^{\mathrm{ref}}$ (here chosen as the median of the observed exposure distribution):

$$
RR_v(x;\mathcal{L})
=
\exp\!\left(
\eta_{i,\mathcal{L}}^{(v)}(x)
-
\eta_{i,\mathcal{L}}^{(v)}(x_v^{\mathrm{ref}})
\right).
$$

For ease of interpretation, relative risk is often expressed as a percentage change:

$$
\%\Delta_v(x;\mathcal{L})
=
\left[ RR_v(x;\mathcal{L}) - 1 \right] \times 100.
$$

#### Temperature

##### Later window (0-40)

```{r}
var <- "tmax"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

prefix <- paste0("cb_", var, "_")
idx <- grepl(prefix, names(cf))

beta   <- cf[idx]
vc_sub <- vc[idx, idx]

# valores para curva completa
x_all   <- wx_use[[var]]
grid    <- sort(unique(as.numeric(quantile(x_all, seq(0.05,0.95,by=0.01)))))

# referência (P50) usada como centragem
P50 <- quantile(x_all, 0.50, na.rm=TRUE)

cp <- crosspred(
  cb,
  coef = beta,
  vcov = vc_sub,
  at   = grid,
  cen  = P50,
  bylag = 1
)

mf <- cp$matfit     # matriz: rows = grid, cols = lags

lag_idx <- suppressWarnings(as.integer(gsub("lag","", colnames(mf))))
if(anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

colsA <- which(lag_idx %in% 0:40)

# efeito cumulativo para CADA temperatura do grid
etaA <- apply(mf[, colsA, drop=FALSE], 1, sum)

RR <- exp(etaA)

rr_sum_t_la <- data.frame(
  tmean = grid,
  RR_med = RR) %>% 
  mutate(
    PERC = (RR_med-1)*100
  )

tmean_RR_second =rr_sum_t_la %>% 
  filter(tmean >=29) %>% 
 #filter(tmean >=5) %>% 
 #filter(tmean >=0.66) %>%
ggplot(aes(x=tmean, y=PERC)) +
  #annotate("rect", xmin = 25, xmax = 26, ymin = -Inf, ymax = Inf,
   #        fill = "lightgreen", color = "lightgreen",alpha = .7) +
  annotate("rect", xmin = 31.5, xmax = 32.5, ymin = -Inf, ymax = Inf, fill = "#E6F2E6", color = "#E6F2E6",alpha = .7) +
  geom_hline(yintercept = 0, linetype=2) +
  geom_vline(xintercept = 31.5, linetype=2) +
  geom_vline(xintercept = 32.5, linetype=2) +
  geom_smooth(size = 3, colour = "#B2182B") +
  labs(x="Temperature (°C)",
       y="Relative risk (%)") +
  theme_bw()+
  theme(
    text = element_text(face = "bold", size = 14) )

tmean_RR_second
```

##### Ealier window (41-70)

```{r}
var <- "tmax"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

prefix <- paste0("cb_", var, "_")
idx <- grepl(prefix, names(cf))

beta   <- cf[idx]
vc_sub <- vc[idx, idx]

# valores para curva completa
x_all   <- wx_use[[var]]
grid    <- sort(unique(as.numeric(quantile(x_all, seq(0.05,0.95,by=0.01)))))

# referência (P50) usada como centragem
P50 <- quantile(x_all, 0.50, na.rm=TRUE)

cp <- crosspred(
  cb,
  coef = beta,
  vcov = vc_sub,
  at   = grid,
  cen  = P50,
  bylag = 1
)

mf <- cp$matfit     # matriz: rows = grid, cols = lags

lag_idx <- suppressWarnings(as.integer(gsub("lag","", colnames(mf))))
if(anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

colsA <- which(lag_idx %in% 41:85)

# efeito cumulativo para CADA temperatura do grid
etaA <- apply(mf[, colsA, drop=FALSE], 1, sum)

RR <- exp(etaA)

rr_sum_t_ea <- data.frame(
  tmean = grid,
  RR_med = RR) %>% 
  mutate(
    PERC = (RR_med-1)*100
  )

tmean_RR_first = rr_sum_t_ea %>% 
  filter(tmean >=29) %>% 
 #filter(tmean >=5) %>% 
 #filter(tmean >=0.66) %>%
ggplot(aes(x=tmean, y=PERC)) +
 #annotate("rect", xmin = 25, xmax = 26, ymin = -Inf, ymax = Inf,
   #        fill = "lightgreen", color = "lightgreen",alpha = .7) +
  annotate("rect", xmin = 32, xmax = 34, ymin = -Inf, ymax = Inf, fill = "#E6F2E6", color = "#E6F2E6",alpha = .7) +
  geom_hline(yintercept = 0, linetype=2) +
  geom_vline(xintercept = 32, linetype=2) +
  geom_vline(xintercept = 34, linetype=2) +
  geom_smooth(size = 3, colour = "#B2182B") +
  labs(x="Temperature (°C)",
       y="Relative risk (%)") +
  theme_bw()+
  theme(
    text = element_text(face = "bold", size = 14),
    axis.title.x = element_blank(),
    axis.text.x = element_blank())


    
tmean_RR_first
```

#### Precipitation

##### Later window (0-40)

```{r}
var <- "rain_cum"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

prefix <- paste0("cb_", var, "_")
idx <- grepl(prefix, names(cf))

beta   <- cf[idx]
vc_sub <- vc[idx, idx]

# valores para curva completa
x_all   <- wx_use[[var]]
grid    <- sort(unique(as.numeric(quantile(x_all, seq(0.05,0.95,by=0.01)))))

# referência (P50) usada como centragem
P50 <- quantile(x_all, 0.50, na.rm=TRUE)

cp <- crosspred(
  cb,
  coef = beta,
  vcov = vc_sub,
  at   = grid,
  cen  = P50,
  bylag = 1
)

mf <- cp$matfit     # matriz: rows = grid, cols = lags

lag_idx <- suppressWarnings(as.integer(gsub("lag","", colnames(mf))))
if(anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

colsA <- which(lag_idx %in% 0:40)

# efeito cumulativo para CADA temperatura do grid
etaA <- apply(mf[, colsA, drop=FALSE], 1, sum)

RR <- exp(etaA)

rr_sum_r_la <- data.frame(
  tmean = grid,
  RR_med = RR) %>% 
  mutate(
    PERC = (RR_med-1)*100
  )

rain_RR_second =rr_sum_r_la %>% 
  #filter(tmean >=25) %>% 
 filter(tmean >=5) %>% 
 #filter(tmean >=0.66) %>%
ggplot(aes(x=tmean, y=PERC)) +
 #annotate("rect", xmin = 5, xmax = 10, ymin = -Inf, ymax = Inf,
  #         fill = "lightgreen", color = "lightgreen",alpha = .7) +
  annotate("rect", xmin = 300, xmax = 600, ymin = -Inf, ymax = Inf, fill = "#E6F2E6", color = "#E6F2E6",alpha = .7) +
  geom_hline(yintercept = 0, linetype=2) +
  geom_vline(xintercept = 300, linetype=2) +
  geom_vline(xintercept = 600, linetype=2) +
  geom_smooth(size = 3, colour = "#2166AC") +
  labs(x="Precipitation (mm)",
       y="") +
  theme_bw()+
  theme(
    text = element_text(face = "bold", size = 14) )

rain_RR_second

```

##### Ealier window (41-70)

```{r}
var <- "rain_cum"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

prefix <- paste0("cb_", var, "_")
idx <- grepl(prefix, names(cf))

beta   <- cf[idx]
vc_sub <- vc[idx, idx]

# valores para curva completa
x_all   <- wx_use[[var]]
grid    <- sort(unique(as.numeric(quantile(x_all, seq(0.05,0.95,by=0.01)))))

# referência (P50) usada como centragem
P50 <- quantile(x_all, 0.50, na.rm=TRUE)

cp <- crosspred(
  cb,
  coef = beta,
  vcov = vc_sub,
  at   = grid,
  cen  = P50,
  bylag = 1
)

mf <- cp$matfit     # matriz: rows = grid, cols = lags

lag_idx <- suppressWarnings(as.integer(gsub("lag","", colnames(mf))))
if(anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

colsA <- which(lag_idx %in% 41:85)

# efeito cumulativo para CADA temperatura do grid
etaA <- apply(mf[, colsA, drop=FALSE], 1, sum)

RR <- exp(etaA)

rr_sum_r_ea <- data.frame(
  tmean = grid,
  RR_med = RR) %>% 
  mutate(
    PERC = (RR_med-1)*100
  )

rain_RR_first = rr_sum_r_ea %>% 
  #filter(tmean >=25) %>% 
 filter(tmean >=5) %>% 
 #filter(tmean >=0.66) %>%
ggplot(aes(x=tmean, y=PERC)) +
  ##annotate("rect", xmin = 5, xmax = 10, ymin = -Inf, ymax = Inf,
    #       fill = "lightgreen", color = "lightgreen",alpha = .7) +
   annotate("rect", xmin = 200, xmax = 350, ymin = -Inf, ymax = Inf, fill = "#E6F2E6", color = "#E6F2E6",alpha = .7) +
  geom_hline(yintercept = 0, linetype=2) +
  geom_vline(xintercept = 200, linetype=2) +
  geom_vline(xintercept = 350, linetype=2) +
  geom_smooth(size = 3, colour = "#2166AC") +
  labs(x="Precipitation (mm)",
       y="") +
  theme_bw()+
  theme(
    text = element_text(face = "bold", size = 14),
    axis.title.x = element_blank(),
    axis.text.x = element_blank())

rain_RR_first
```

#### VPD

##### Later window (0-40)

```{r}
var <- "vpd"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

prefix <- paste0("cb_", var, "_")
idx <- grepl(prefix, names(cf))

beta   <- cf[idx]
vc_sub <- vc[idx, idx]

# valores para curva completa
x_all   <- wx_use[[var]]
grid    <- sort(unique(as.numeric(quantile(x_all, seq(0.05,0.95,by=0.01)))))

# referência (P50) usada como centragem
P50 <- quantile(x_all, 0.50, na.rm=TRUE)

cp <- crosspred(
  cb,
  coef = beta,
  vcov = vc_sub,
  at   = grid,
  cen  = P50,
  bylag = 1
)

mf <- cp$matfit     # matriz: rows = grid, cols = lags

lag_idx <- suppressWarnings(as.integer(gsub("lag","", colnames(mf))))
if(anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

colsA <- which(lag_idx %in% 0:40)

# efeito cumulativo para CADA temperatura do grid
etaA <- apply(mf[, colsA, drop=FALSE], 1, sum)

RR <- exp(etaA)

rr_sum_v_la <- data.frame(
  tmean = grid,
  RR_med = RR) %>% 
  mutate(
    PERC = (RR_med-1)*100
  )

rr_sum_v_la$phase = "Later"

vpd_RR_second =rr_sum_v_la %>% 
  #filter(tmean >=25) %>% 
 #filter(tmean >=5) %>% 
 filter(tmean >=0.66) %>%
 # filter(tmean >=81) %>%
ggplot(aes(x=tmean, y=PERC)) +
##annotate("rect", xmin = 0.66, xmax = 1, ymin = -Inf, ymax = Inf,
  ##         fill = "lightgreen", color = "lightgreen",alpha = .7) +
   annotate("rect", xmin = 1.3, xmax = 1.5, ymin = -Inf, ymax = Inf, fill = "#E6F2E6", color = "#E6F2E6",alpha = .7) +
  geom_hline(yintercept = 0, linetype=2) +
  geom_vline(xintercept = 1.3, linetype=2) +
  geom_vline(xintercept = 1.5, linetype=2) +
  geom_smooth(size = 3, colour = "green4") +
  labs(x="VPD (kPa)",
       y="") +
  facet_wrap(~phase, strip.position = "right")+
  theme_bw()+
  theme(
    text = element_text(face = "bold", size = 14),
    strip.background = element_blank(),
    strip.text = element_text(size = 18, face = "bold"))

vpd_RR_second

```

##### Ealier window (41-70)

```{r}
var <- "vpd"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

prefix <- paste0("cb_", var, "_")
idx <- grepl(prefix, names(cf))

beta   <- cf[idx]
vc_sub <- vc[idx, idx]

# valores para curva completa
x_all   <- wx_use[[var]]
grid    <- sort(unique(as.numeric(quantile(x_all, seq(0.05,0.95,by=0.01)))))

# referência (P50) usada como centragem
P50 <- quantile(x_all, 0.50, na.rm=TRUE)

cp <- crosspred(
  cb,
  coef = beta,
  vcov = vc_sub,
  at   = grid,
  cen  = P50,
  bylag = 1
)

mf <- cp$matfit     # matriz: rows = grid, cols = lags

lag_idx <- suppressWarnings(as.integer(gsub("lag","", colnames(mf))))
if(anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

colsA <- which(lag_idx %in% 41:85)

# efeito cumulativo para CADA temperatura do grid
etaA <- apply(mf[, colsA, drop=FALSE], 1, sum)

RR <- exp(etaA)

rr_sum_v_ea <- data.frame(
  tmean = grid,
  RR_med = RR) %>% 
  mutate(
    PERC = (RR_med-1)*100
  )

rr_sum_v_ea$phase = "Ealier"


vpd_RR_first = rr_sum_v_ea %>% 
 # filter(tmean >=25) %>% 
 #filter(tmean >=5) %>% 
 filter(tmean >=0.66) %>%
ggplot(aes(x=tmean, y=PERC)) +
##annotate("rect", xmin = 0.66, xmax = 1, ymin = -Inf, ymax = Inf,
  #         fill = "lightgreen", color = "lightgreen", alpha = .7) +
  annotate("rect", xmin = 0.66, xmax = 0.80, ymin = -Inf, ymax = Inf, fill = "#E6F2E6", color = "#E6F2E6",alpha = .7) +
  geom_hline(yintercept = 0, linetype=2) +
  geom_vline(xintercept = 0.66, linetype=2) +
  geom_vline(xintercept = 0.80, linetype=2) +
  geom_smooth(size = 3, colour = "green4") +
  labs(x="Precipitation (mm)",
       y="") +
    facet_wrap(~phase, strip.position = "right")+
  theme_bw()+
  theme(
    text = element_text(face = "bold", size = 14),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(size = 18, face = "bold"))

vpd_RR_first
```


```{r}
library(patchwork)

tmean_RR_first/tmean_RR_second|rain_RR_first/rain_RR_second|vpd_RR_first/vpd_RR_second


ggsave("fig/RR1.png", dpi = 600, width = 12, height = 6)
```

### Lag movel

#### Temperature

```{r}
library(dlnm)
library(dplyr)
library(ggplot2)
library(tidyr)

# ==============================================================
# 1) EXTRAIR COEFICIENTES
# ==============================================================

var <- "tmax"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

idx     <- grepl("^cb_tmax_", names(cf))
beta    <- cf[idx]
vc_sub  <- vc[idx, idx]

stopifnot(length(beta) == ncol(cb))
stopifnot(nrow(vc_sub) == ncol(cb))

# ==============================================================
# 2) DEFINIR GRID E FAIXA DE TMEAN PARA O HEATMAP
# ==============================================================

x_all <- wx_use[[var]]
P50   <- quantile(x_all, 0.50, na.rm = TRUE)

# grid usado pelo DLNM
grid <- sort(unique(as.numeric(quantile(
  x_all, seq(0.05, 0.95, by = 0.01), na.rm = TRUE
))))

# valores desejados para o heatmap (26 a 30 °C)
#at_vals <- seq(1, 2, length.out = 10000)
at_vals <- seq(29, 34, length.out = 10000)

# ajustar ao range REAL do grid (evita extrapolação)
at_vals <- at_vals[at_vals >= min(grid) & at_vals <= max(grid)]

# ==============================================================
# 3) crosspred EM TODOS OS at_vals (lag‑móvel)
# ==============================================================

cp <- crosspred(
  cb,
  coef  = beta,
  vcov  = vc_sub,
  at    = at_vals,
  cen   = P50,
  bylag = 1
)

# matriz: linhas = valores de tmean; colunas = lags
mf <- cp$matfit

# criar vetor de lags
lag_idx <- suppressWarnings(as.integer(gsub("lag","",colnames(mf))))
if (anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

# ==============================================================
# 4) Converter para RR(lag) e depois % (% change)
# ==============================================================

RR_mat   <- exp(mf)
PERC_mat <- (RR_mat - 1) * 100

# pivô para formato tidy
df_heat <- as.data.frame(PERC_mat)
colnames(df_heat) <- paste0("lag_", lag_idx)
df_heat <- df_heat %>%
  mutate(tmean_val = at_vals) %>%
  pivot_longer(cols = starts_with("lag_"),
               names_to = "lag",
               values_to = "PERC") %>%
  mutate(lag = as.integer(gsub("lag_", "", lag)))

```

```{r}
library(ggplot2)
library(dplyr)
library(scales)
library(metR)

# 2a) (opcional mas muito útil) "aparar" outliers para a escala de cor
#     e forçar limites simétricos ao redor de zero:
lims <- quantile(df_heat$PERC, c(0.02, 0.98), na.rm = TRUE)
max_abs <- max(abs(lims))
df_plot <- df_heat %>%
  mutate(PERC_c = pmax(pmin(PERC,  max_abs), -max_abs))  # winsoriza coloração

# 2b) Heatmap + isolinhas
tmean_movel = df_plot %>% 
ggplot(aes(x = lag, y = tmean_val, fill = PERC_c)) +
  # raster interpolado fica mais liso na tela
  geom_raster(interpolate = TRUE) +
  # isolinhas (0% e ±5, ±10, ±15...) – ajuste 'breaks' conforme seus dados
  geom_contour(aes(z = PERC), color = "black", size = 0.25,
               breaks = c(-15, -10, -5, 0, 5, 10, 15)) +
  # rótulo só na isolinha zero (neutro) para guiar leitura
  geom_text_contour(aes(z = PERC), breaks = 0, stroke = 0.15,
                    label.padding = unit(0.15, "lines"), size = 3.2) +
  # paleta perceptual e centrada no 0 (diverging)
  scale_fill_gradient2(
    low = "#2c7bb6", mid = "white", high = "#d7191c",
    midpoint = 0, limits = c(-max_abs, max_abs),
    oob = squish, name = "RR (%)"
  ) +
  # eixos e guias
  scale_x_continuous(breaks = pretty(df_plot$lag, n = 8), expand = c(0, 0)) +
  scale_y_continuous(breaks = pretty(df_plot$tmean_val, n = 6), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  labs(
    x = "Lag (days before rating)",
    y = "Temperature (°C)") +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold", size = 12),
    axis.text  = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10)
  )

tmean_movel
```

##### Severity

```{r}
cb_cols <- names(dat)[grepl("^cb_(tmax|vpd|rain_cum)_", names(dat))]
fml_sel <- as.formula(paste("sev ~ 1 + (1|siteyear) +", paste(cb_cols, collapse = " + ")))

#dat$sev  = dat$sev/100

fit_sel <- glmmTMB::glmmTMB(
  fml_sel, data = dat,
  family = glmmTMB::beta_family(link = "logit")
)

summary(fit_sel)
```

```{r}

# Pacotes
library(dlnm)
library(glmmTMB)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
# Se quiser isolinhas com rótulo: remotes::install_github("teunbrand/ggtextpath")
# library(ggtextpath) # para geom_textcontour (opcional)

# --------------------------------------------------------------
# Entradas esperadas já existentes no seu ambiente:
# - fit_sel         : modelo glmmTMB ajustado com cb_*(tmean, vpd, rain, ...)
# - cb_templates    : lista com os crossbasis (p.ex., cb_templates[["tmean"]])
# - wx_use          : lista com séries originais (p.ex., wx_use[["tmean"]])
# - LAG_MAX         : lag máximo usado ao construir os crossbasis (>= 70)
# --------------------------------------------------------------

# ------------ 1) Parâmetros do heatmap -------------
var      <- "tmax"                     # variável-alvo
cb       <- cb_templates[[var]]         # crossbasis correspondente
x_all    <- wx_use[[var]]
P50      <- as.numeric(quantile(x_all, 0.50, na.rm = TRUE))

# Faixa para o heatmap (26–30 °C) aparada ao grid real (evita extrapolação)
grid_obs <- sort(unique(as.numeric(quantile(x_all, seq(0.05, 0.95, by=0.01), na.rm=TRUE))))
at_vals  <- seq(29, 34, length.out = 1000)
at_vals  <- at_vals[at_vals >= min(grid_obs) & at_vals <= max(grid_obs)]

# ------------ 2) Extrair coeficientes/VCOV do bloco DLNM de tmean -------------
cf <- coef(summary(fit_sel))$cond[, 1]
vc <- vcov(fit_sel)$cond

pref     <- paste0("^cb_", var, "_")
idx_beta <- grep(pref, names(cf))
stopifnot(length(idx_beta) > 0)
beta     <- cf[idx_beta]

idx_rows <- grep(pref, rownames(vc))
idx_cols <- grep(pref, colnames(vc))
V        <- vc[idx_rows, idx_cols, drop = FALSE]

# Checagem de consistência com o crossbasis
stopifnot(length(beta) == ncol(cb))
stopifnot(nrow(V) == ncol(cb), ncol(V) == ncol(cb))

# ------------ 3) crosspred em todos os at_vals, lag móvel -------------
cp <- dlnm::crosspred(
  cb,
  coef  = beta,
  vcov  = V,
  at    = at_vals,
  cen   = P50,
  bylag = 1
)

# Matriz no preditor linear: linhas = tmean (at_vals), colunas = lags
mf <- cp$matfit

# Recuperar e ordenar os índices de lag
lag_idx <- suppressWarnings(as.integer(gsub("lag", "", colnames(mf))))
if (anyNA(lag_idx)) {
  lag_idx <- 0:(ncol(mf)-1)
  colnames(mf) <- paste0("lag", lag_idx)
}
ord <- order(lag_idx)
mf  <- mf[, ord, drop = FALSE]
lag_idx <- lag_idx[ord]

# Limitar à faixa de interesse (0–70 dias)
lag_max_plot <- min(85, max(lag_idx))
keep_cols    <- which(lag_idx >= 0 & lag_idx <= lag_max_plot)
mf           <- mf[, keep_cols, drop = FALSE]
lag_idx      <- lag_idx[keep_cols]

# ------------ 4) Converter Δη acumulado (0→L) em sev final -------------
# Baseline: mesmo critério do seu Script 2 (média das predições)
# >>> Se quiser apenas efeitos fixos, troque pela linha comentada.
sev_base <- mean(predict(fit_sel, type = "response"))
# sev_base <- mean(predict(fit_sel, type = "response", re.form = NA))  # apenas FE

# cumsum ao longo do eixo de lag produz o acúmulo 0→L para cada tmean
mf_cum <- t(apply(mf, 1, cumsum))  # dimensões: length(at_vals) x length(lag_idx)

# Mapear para severidade final (escala 0–1), depois para %
logit <- function(p) qlogis(p)
ilogit <- function(x) plogis(x)

SEV_mat <- ilogit(logit(sev_base) + mf_cum) # (0–1)
SEVpct_mat <- 100 * SEV_mat                 # (0–100%)

# ------------ 5) Data frame "long" para o heatmap -------------
df_sev <- as.data.frame(SEVpct_mat)
colnames(df_sev) <- paste0("lag_", lag_idx)
df_sev <- df_sev |>
  mutate(tmean_val = at_vals) |>
  pivot_longer(
    cols = starts_with("lag_"),
    names_to = "lag",
    values_to = "SEV_pct"
  ) |>
  mutate(lag = as.integer(gsub("lag_", "", lag)))


# --------------------------------------------------------------
# (OPCIONAL) 7) Versão por janelas FIXAS (p.ex., 0–40 e 41–70)
# Produz uma curva de SEV_final vs tmean para cada janela
# --------------------------------------------------------------


make_band_curve <- function(mf_row_by_lag, lag_idx, band) {
  cols <- which(lag_idx %in% band)
  if (length(cols) == 0) return(NA_real_)
  sum(mf_row_by_lag[cols], na.rm = TRUE)
}

bands <- list(`Later` = 0:40, `Earlier` = 41:85)
eta_band <- lapply(bands, function(bd) {
  apply(mf, 1, make_band_curve, lag_idx = lag_idx, band = bd)
})
band_df <- bind_rows(lapply(names(eta_band), function(nm) {
  tibble(
    tmean_val = at_vals,
    band      = nm,
    SEV_pct   = 100 * ilogit(logit(sev_base) + eta_band[[nm]])
  )
}), .id = NULL)


sev_tmean = ggplot(band_df, aes(tmean_val, SEV_pct, linetype = band)) +
  #geom_hline(yintercept = 100*sev_base, linetype = 2, alpha = 0.6) +
  geom_smooth(size = 2, se = F, color = "#B2182B") +
  #scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Temperature (°C)",
    y = "Severity (%)",
    linetype = "Phases") +
  theme_bw(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text  = element_text(size = 10))


sev_tmean
```

#### Precipitation

```{r}
library(dlnm)
library(dplyr)
library(ggplot2)
library(tidyr)

# ==============================================================
# 1) EXTRAIR COEFICIENTES
# ==============================================================

var <- "rain_cum"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

idx     <- grepl("^cb_rain_cum_", names(cf))
beta    <- cf[idx]
vc_sub  <- vc[idx, idx]

stopifnot(length(beta) == ncol(cb))
stopifnot(nrow(vc_sub) == ncol(cb))

# ==============================================================
# 2) DEFINIR GRID E FAIXA DE TMEAN PARA O HEATMAP
# ==============================================================

x_all <- wx_use[[var]]
P50   <- quantile(x_all, 0.50, na.rm = TRUE)

# grid usado pelo DLNM
grid <- sort(unique(as.numeric(quantile(
  x_all, seq(0.05, 0.95, by = 0.01), na.rm = TRUE
))))

# valores desejados para o heatmap (26 a 30 °C)
#at_vals <- seq(1, 2, length.out = 10000)
at_vals <- seq(0, 600, length.out = 10000)

# ajustar ao range REAL do grid (evita extrapolação)
at_vals <- at_vals[at_vals >= min(grid) & at_vals <= max(grid)]

# ==============================================================
# 3) crosspred EM TODOS OS at_vals (lag‑móvel)
# ==============================================================

cp <- crosspred(
  cb,
  coef  = beta,
  vcov  = vc_sub,
  at    = at_vals,
  cen   = P50,
  bylag = 1
)

# matriz: linhas = valores de tmean; colunas = lags
mf <- cp$matfit

# criar vetor de lags
lag_idx <- suppressWarnings(as.integer(gsub("lag","",colnames(mf))))
if (anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

# ==============================================================
# 4) Converter para RR(lag) e depois % (% change)
# ==============================================================

RR_mat   <- exp(mf)
PERC_mat <- (RR_mat - 1) * 100

# pivô para formato tidy
df_heat <- as.data.frame(PERC_mat)
colnames(df_heat) <- paste0("lag_", lag_idx)
df_heat <- df_heat %>%
  mutate(tmean_val = at_vals) %>%
  pivot_longer(cols = starts_with("lag_"),
               names_to = "lag",
               values_to = "PERC") %>%
  mutate(lag = as.integer(gsub("lag_", "", lag)))

```

```{r}
library(ggplot2)
library(dplyr)
library(scales)

# 2a) (opcional mas muito útil) "aparar" outliers para a escala de cor
#     e forçar limites simétricos ao redor de zero:
lims <- quantile(df_heat$PERC, c(0.02, 0.98), na.rm = TRUE)
max_abs <- max(abs(lims))
df_plot <- df_heat %>%
  mutate(PERC_c = pmax(pmin(PERC,  max_abs), -max_abs))  # winsoriza coloração

# 2b) Heatmap + isolinhas
rain_movel = df_plot %>% 
ggplot(aes(x = lag, y = tmean_val, fill = PERC_c)) +
  # raster interpolado fica mais liso na tela
  geom_raster(interpolate = TRUE) +
  # isolinhas (0% e ±5, ±10, ±15...) – ajuste 'breaks' conforme seus dados
  geom_contour(aes(z = PERC), color = "black", size = 0.25) +
  # rótulo só na isolinha zero (neutro) para guiar leitura
  geom_text_contour(aes(z = PERC), breaks = 0, stroke = 0.15,
                    label.padding = unit(0.15, "lines"), size = 3.2) +
  # paleta perceptual e centrada no 0 (diverging)
  scale_fill_gradient2(
    low = "#2c7bb6", mid = "white", high = "#d7191c",
    midpoint = 0, limits = c(-max_abs, max_abs),
    oob = squish, name = "RR (%)"
  ) +
  # eixos e guias
  scale_x_continuous(breaks = pretty(df_plot$lag, n = 8), expand = c(0, 0)) +
  scale_y_continuous(breaks = pretty(df_plot$tmean_val, n = 6), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  labs(
    x = "Lag (days before rating)",
    y = "Precipitation (mm)") +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold", size = 12),
    axis.text  = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10)
  )

rain_movel

```

##### Severity

```{r}

# Pacotes
library(dlnm)
library(glmmTMB)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
# Se quiser isolinhas com rótulo: remotes::install_github("teunbrand/ggtextpath")
# library(ggtextpath) # para geom_textcontour (opcional)

# --------------------------------------------------------------
# Entradas esperadas já existentes no seu ambiente:
# - fit_sel         : modelo glmmTMB ajustado com cb_*(tmean, vpd, rain, ...)
# - cb_templates    : lista com os crossbasis (p.ex., cb_templates[["tmean"]])
# - wx_use          : lista com séries originais (p.ex., wx_use[["tmean"]])
# - LAG_MAX         : lag máximo usado ao construir os crossbasis (>= 70)
# --------------------------------------------------------------

# ------------ 1) Parâmetros do heatmap -------------
var      <- "rain_cum"                     # variável-alvo
cb       <- cb_templates[[var]]         # crossbasis correspondente
x_all    <- wx_use[[var]]
P50      <- as.numeric(quantile(x_all, 0.50, na.rm = TRUE))

# Faixa para o heatmap (26–30 °C) aparada ao grid real (evita extrapolação)
grid_obs <- sort(unique(as.numeric(quantile(x_all, seq(0.05, 0.95, by=0.01), na.rm=TRUE))))
at_vals  <- seq(0, 600, length.out = 1000)
at_vals  <- at_vals[at_vals >= min(grid_obs) & at_vals <= max(grid_obs)]

# ------------ 2) Extrair coeficientes/VCOV do bloco DLNM de tmean -------------
cf <- coef(summary(fit_sel))$cond[, 1]
vc <- vcov(fit_sel)$cond

pref     <- paste0("^cb_", var, "_")
idx_beta <- grep(pref, names(cf))
stopifnot(length(idx_beta) > 0)
beta     <- cf[idx_beta]

idx_rows <- grep(pref, rownames(vc))
idx_cols <- grep(pref, colnames(vc))
V        <- vc[idx_rows, idx_cols, drop = FALSE]

# Checagem de consistência com o crossbasis
stopifnot(length(beta) == ncol(cb))
stopifnot(nrow(V) == ncol(cb), ncol(V) == ncol(cb))

# ------------ 3) crosspred em todos os at_vals, lag móvel -------------
cp <- dlnm::crosspred(
  cb,
  coef  = beta,
  vcov  = V,
  at    = at_vals,
  cen   = P50,
  bylag = 1
)

# Matriz no preditor linear: linhas = tmean (at_vals), colunas = lags
mf <- cp$matfit

# Recuperar e ordenar os índices de lag
lag_idx <- suppressWarnings(as.integer(gsub("lag", "", colnames(mf))))
if (anyNA(lag_idx)) {
  lag_idx <- 0:(ncol(mf)-1)
  colnames(mf) <- paste0("lag", lag_idx)
}
ord <- order(lag_idx)
mf  <- mf[, ord, drop = FALSE]
lag_idx <- lag_idx[ord]

# Limitar à faixa de interesse (0–70 dias)
lag_max_plot <- min(85, max(lag_idx))
keep_cols    <- which(lag_idx >= 0 & lag_idx <= lag_max_plot)
mf           <- mf[, keep_cols, drop = FALSE]
lag_idx      <- lag_idx[keep_cols]

# ------------ 4) Converter Δη acumulado (0→L) em sev final -------------
# Baseline: mesmo critério do seu Script 2 (média das predições)
# >>> Se quiser apenas efeitos fixos, troque pela linha comentada.
sev_base <- mean(predict(fit_sel, type = "response"))
# sev_base <- mean(predict(fit_sel, type = "response", re.form = NA))  # apenas FE

# cumsum ao longo do eixo de lag produz o acúmulo 0→L para cada tmean
mf_cum <- t(apply(mf, 1, cumsum))  # dimensões: length(at_vals) x length(lag_idx)

# Mapear para severidade final (escala 0–1), depois para %
logit <- function(p) qlogis(p)
ilogit <- function(x) plogis(x)

SEV_mat <- ilogit(logit(sev_base) + mf_cum) # (0–1)
SEVpct_mat <- 100 * SEV_mat                 # (0–100%)

# ------------ 5) Data frame "long" para o heatmap -------------
df_sev <- as.data.frame(SEVpct_mat)
colnames(df_sev) <- paste0("lag_", lag_idx)
df_sev <- df_sev |>
  mutate(tmean_val = at_vals) |>
  pivot_longer(
    cols = starts_with("lag_"),
    names_to = "lag",
    values_to = "SEV_pct"
  ) |>
  mutate(lag = as.integer(gsub("lag_", "", lag)))


# --------------------------------------------------------------
# (OPCIONAL) 7) Versão por janelas FIXAS (p.ex., 0–40 e 41–70)
# Produz uma curva de SEV_final vs tmean para cada janela
# --------------------------------------------------------------


make_band_curve <- function(mf_row_by_lag, lag_idx, band) {
  cols <- which(lag_idx %in% band)
  if (length(cols) == 0) return(NA_real_)
  sum(mf_row_by_lag[cols], na.rm = TRUE)
}

bands <- list(`Later` = 0:40, `Earlier` = 41:85)
eta_band <- lapply(bands, function(bd) {
  apply(mf, 1, make_band_curve, lag_idx = lag_idx, band = bd)
})
band_df <- bind_rows(lapply(names(eta_band), function(nm) {
  tibble(
    tmean_val = at_vals,
    band      = nm,
    SEV_pct   = 100 * ilogit(logit(sev_base) + eta_band[[nm]])
  )
}), .id = NULL)

sev_rain = ggplot(band_df, aes(tmean_val, SEV_pct, linetype = band)) +
  #geom_hline(yintercept = 100*sev_base, linetype = 2, alpha = 0.6) +
  geom_smooth(size = 2, se = F, colour = "#2166AC") +
  #scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Precipitation (mm)",
    y = "Severity (%)",
    linetype = "Phases") +
  theme_bw(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text  = element_text(size = 10))

sev_rain
```

#### VPD

```{r}
library(dlnm)
library(dplyr)
library(ggplot2)
library(tidyr)

# ==============================================================
# 1) EXTRAIR COEFICIENTES
# ==============================================================

var <- "vpd"
cb  <- cb_templates[[var]]

cf  <- coef(summary(fit_tmb))$cond[,1]
vc  <- vcov(fit_tmb)$cond

idx     <- grepl("^cb_vpd_", names(cf))
beta    <- cf[idx]
vc_sub  <- vc[idx, idx]

stopifnot(length(beta) == ncol(cb))
stopifnot(nrow(vc_sub) == ncol(cb))

# ==============================================================
# 2) DEFINIR GRID E FAIXA DE TMEAN PARA O HEATMAP
# ==============================================================

x_all <- wx_use[[var]]
P50   <- quantile(x_all, 0.50, na.rm = TRUE)

# grid usado pelo DLNM
grid <- sort(unique(as.numeric(quantile(
  x_all, seq(0.05, 0.95, by = 0.01), na.rm = TRUE
))))

# valores desejados para o heatmap (26 a 30 °C)
#at_vals <- seq(1, 2, length.out = 10000)
at_vals <- seq(0.66, 1.5, length.out = 10000)

# ajustar ao range REAL do grid (evita extrapolação)
at_vals <- at_vals[at_vals >= min(grid) & at_vals <= max(grid)]

# ==============================================================
# 3) crosspred EM TODOS OS at_vals (lag‑móvel)
# ==============================================================

cp <- crosspred(
  cb,
  coef  = beta,
  vcov  = vc_sub,
  at    = at_vals,
  cen   = P50,
  bylag = 1
)

# matriz: linhas = valores de tmean; colunas = lags
mf <- cp$matfit

# criar vetor de lags
lag_idx <- suppressWarnings(as.integer(gsub("lag","",colnames(mf))))
if (anyNA(lag_idx)) lag_idx <- 0:(ncol(mf)-1)

# ==============================================================
# 4) Converter para RR(lag) e depois % (% change)
# ==============================================================

RR_mat   <- exp(mf)
PERC_mat <- (RR_mat - 1) * 100

# pivô para formato tidy
df_heat <- as.data.frame(PERC_mat)
colnames(df_heat) <- paste0("lag_", lag_idx)
df_heat <- df_heat %>%
  mutate(tmean_val = at_vals) %>%
  pivot_longer(cols = starts_with("lag_"),
               names_to = "lag",
               values_to = "PERC") %>%
  mutate(lag = as.integer(gsub("lag_", "", lag)))

```

```{r}
library(ggplot2)
library(dplyr)
library(scales)

# 2a) (opcional mas muito útil) "aparar" outliers para a escala de cor
#     e forçar limites simétricos ao redor de zero:
lims <- quantile(df_heat$PERC, c(0.02, 0.98), na.rm = TRUE)
max_abs <- max(abs(lims))
df_plot <- df_heat %>%
  mutate(PERC_c = pmax(pmin(PERC,  max_abs), -max_abs))  # winsoriza coloração

# 2b) Heatmap + isolinhas
vpd_movel = df_plot %>% 
ggplot(aes(x = lag, y = tmean_val, fill = PERC_c)) +
  # raster interpolado fica mais liso na tela
  geom_raster(interpolate = TRUE) +
  # isolinhas (0% e ±5, ±10, ±15...) – ajuste 'breaks' conforme seus dados
  geom_contour(aes(z = PERC), color = "black", size = 0.25,
               breaks = c(-15, -10, -5, 0, 5, 10, 15)) +
  # rótulo só na isolinha zero (neutro) para guiar leitura
  geom_text_contour(aes(z = PERC), breaks = 0, stroke = 0.15,
                    label.padding = unit(0.15, "lines"), size = 3.2) +
  # paleta perceptual e centrada no 0 (diverging)
  scale_fill_gradient2(
    low = "#2c7bb6", mid = "white", high = "#d7191c",
    midpoint = 0, limits = c(-max_abs, max_abs),
    oob = squish, name = "RR (%)"
  ) +
  # eixos e guias
  scale_x_continuous(breaks = pretty(df_plot$lag, n = 8), expand = c(0, 0)) +
  scale_y_continuous(breaks = pretty(df_plot$tmean_val, n = 6), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  labs(
    x = "Lag (days before rating)",
    y = "VPD (kPa)") +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold", size = 12),
    axis.text  = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10)
  )

vpd_movel
```
##### Severity

```{r}

# Pacotes
library(dlnm)
library(glmmTMB)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
# Se quiser isolinhas com rótulo: remotes::install_github("teunbrand/ggtextpath")
# library(ggtextpath) # para geom_textcontour (opcional)

# --------------------------------------------------------------
# Entradas esperadas já existentes no seu ambiente:
# - fit_sel         : modelo glmmTMB ajustado com cb_*(tmean, vpd, rain, ...)
# - cb_templates    : lista com os crossbasis (p.ex., cb_templates[["tmean"]])
# - wx_use          : lista com séries originais (p.ex., wx_use[["tmean"]])
# - LAG_MAX         : lag máximo usado ao construir os crossbasis (>= 70)
# --------------------------------------------------------------

# ------------ 1) Parâmetros do heatmap -------------
var      <- "vpd"                     # variável-alvo
cb       <- cb_templates[[var]]         # crossbasis correspondente
x_all    <- wx_use[[var]]
P50      <- as.numeric(quantile(x_all, 0.50, na.rm = TRUE))

# Faixa para o heatmap (26–30 °C) aparada ao grid real (evita extrapolação)
grid_obs <- sort(unique(as.numeric(quantile(x_all, seq(0.05, 0.95, by=0.01), na.rm=TRUE))))
at_vals  <- seq(0.66, 1.5, length.out = 1000)
at_vals  <- at_vals[at_vals >= min(grid_obs) & at_vals <= max(grid_obs)]

# ------------ 2) Extrair coeficientes/VCOV do bloco DLNM de tmean -------------
cf <- coef(summary(fit_sel))$cond[, 1]
vc <- vcov(fit_sel)$cond

pref     <- paste0("^cb_", var, "_")
idx_beta <- grep(pref, names(cf))
stopifnot(length(idx_beta) > 0)
beta     <- cf[idx_beta]

idx_rows <- grep(pref, rownames(vc))
idx_cols <- grep(pref, colnames(vc))
V        <- vc[idx_rows, idx_cols, drop = FALSE]

# Checagem de consistência com o crossbasis
stopifnot(length(beta) == ncol(cb))
stopifnot(nrow(V) == ncol(cb), ncol(V) == ncol(cb))

# ------------ 3) crosspred em todos os at_vals, lag móvel -------------
cp <- dlnm::crosspred(
  cb,
  coef  = beta,
  vcov  = V,
  at    = at_vals,
  cen   = P50,
  bylag = 1
)

# Matriz no preditor linear: linhas = tmean (at_vals), colunas = lags
mf <- cp$matfit

# Recuperar e ordenar os índices de lag
lag_idx <- suppressWarnings(as.integer(gsub("lag", "", colnames(mf))))
if (anyNA(lag_idx)) {
  lag_idx <- 0:(ncol(mf)-1)
  colnames(mf) <- paste0("lag", lag_idx)
}
ord <- order(lag_idx)
mf  <- mf[, ord, drop = FALSE]
lag_idx <- lag_idx[ord]

# Limitar à faixa de interesse (0–70 dias)
lag_max_plot <- min(85, max(lag_idx))
keep_cols    <- which(lag_idx >= 0 & lag_idx <= lag_max_plot)
mf           <- mf[, keep_cols, drop = FALSE]
lag_idx      <- lag_idx[keep_cols]

# ------------ 4) Converter Δη acumulado (0→L) em sev final -------------
# Baseline: mesmo critério do seu Script 2 (média das predições)
# >>> Se quiser apenas efeitos fixos, troque pela linha comentada.
sev_base <- mean(predict(fit_sel, type = "response"))
# sev_base <- mean(predict(fit_sel, type = "response", re.form = NA))  # apenas FE

# cumsum ao longo do eixo de lag produz o acúmulo 0→L para cada tmean
mf_cum <- t(apply(mf, 1, cumsum))  # dimensões: length(at_vals) x length(lag_idx)

# Mapear para severidade final (escala 0–1), depois para %
logit <- function(p) qlogis(p)
ilogit <- function(x) plogis(x)

SEV_mat <- ilogit(logit(sev_base) + mf_cum) # (0–1)
SEVpct_mat <- 100 * SEV_mat                 # (0–100%)

# ------------ 5) Data frame "long" para o heatmap -------------
df_sev <- as.data.frame(SEVpct_mat)
colnames(df_sev) <- paste0("lag_", lag_idx)
df_sev <- df_sev |>
  mutate(tmean_val = at_vals) |>
  pivot_longer(
    cols = starts_with("lag_"),
    names_to = "lag",
    values_to = "SEV_pct"
  ) |>
  mutate(lag = as.integer(gsub("lag_", "", lag)))


# --------------------------------------------------------------
# (OPCIONAL) 7) Versão por janelas FIXAS (p.ex., 0–40 e 41–70)
# Produz uma curva de SEV_final vs tmean para cada janela
# --------------------------------------------------------------


make_band_curve <- function(mf_row_by_lag, lag_idx, band) {
  cols <- which(lag_idx %in% band)
  if (length(cols) == 0) return(NA_real_)
  sum(mf_row_by_lag[cols], na.rm = TRUE)
}

bands <- list(`Later` = 0:40, `Earlier` = 41:85)
eta_band <- lapply(bands, function(bd) {
  apply(mf, 1, make_band_curve, lag_idx = lag_idx, band = bd)
})
band_df <- bind_rows(lapply(names(eta_band), function(nm) {
  tibble(
    tmean_val = at_vals,
    band      = nm,
    SEV_pct   = 100 * ilogit(logit(sev_base) + eta_band[[nm]])
  )
}), .id = NULL)

sev_vpd = ggplot(band_df, aes(tmean_val, SEV_pct, linetype = band)) +
  #geom_hline(yintercept = 100*sev_base, linetype = 2, alpha = 0.6) +
  geom_smooth(size = 2, se = F, colour = "green4") +
  #scale_color_brewer(palette = "Dark2") +
  labs(
    x = "VPD (kPa)",
    y = "Severity (%)",
    linetype = "Phases") +
  theme_bw(base_size = 12) +
  theme(
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text  = element_text(size = 10))

sev_vpd
```

```{r}
tmean_movel/rain_movel/vpd_movel|sev_tmean/sev_rain/sev_vpd

ggsave("fig/RR_movel_severity2.png", dpi = 600, width = 10, height = 6)

#ggsave("fig/RR_movel_severity.png", dpi = 600, width = 12, height = 8)
```


# Discriptive plotting
```{r}
wx_summarise = wx_use %>% 
  filter(dpp <= 85) %>% 
  group_by(dpp) %>% 
  summarise(
    TMEAN = mean(tmean),
    TMEAN_low = quantile(tmean,0.025),
    TMEAN_high = quantile(tmean,0.975),
    RAIN = mean(rain),
    RAIN_low = quantile(rain,0.025),
    RAIN_high = quantile(rain,0.975),
    VPD = mean(vpd),
    VPD_low = quantile(vpd,0.025),
    VPD_high = quantile(vpd,0.975)
  )



library(ggplot2)

tmean_plot = ggplot() +
  geom_smooth(
    data = wx_use %>% filter(dpp <= 85),
    aes(dpp, tmean, group = epi_id),
    color = "#B2182B",
    alpha = 0.1,
    linewidth = 0.6, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, TMEAN),
    color = "black",
    linewidth = 1.2, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, TMEAN_low),
    linetype = "dashed",
    color = "black",
    linewidth = 0.8, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, TMEAN_high),
    linetype = "dashed",
    color = "black",
    linewidth = 0.8, se = F
  ) +
  labs(
    x = "Days",
    y = "Temperature (°C)"
  ) +
  theme_bw()+
  theme(text = element_text(size = 10),
        axis.title = element_text(face = "bold", size = 12))



vpd_plot = ggplot() +
  geom_smooth(
    data = wx_use %>% filter(dpp <= 85),
    aes(dpp, vpd, group = epi_id),
    color = "green4",
    alpha = 0.1,
    linewidth = 0.6, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, VPD),
    color = "black",
    linewidth = 1.2, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, VPD_low),
    linetype = "dashed",
    color = "black",
    linewidth = 0.8, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, VPD_high),
    linetype = "dashed",
    color = "black",
    linewidth = 0.8, se = F
  ) +
  labs(
    x = "Days",
    y = "VPD (kPa)"
  ) +
  theme_bw()+
  theme(text = element_text(size = 10),
        axis.title = element_text(face = "bold", size = 12))



rain_plot = ggplot() +
  geom_smooth(
    data = wx_use %>% filter(dpp <= 85),
    aes(dpp, rain, group = epi_id),
    color = "#2166AC",
    alpha = 0.1,
    linewidth = 0.6, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, RAIN),
    color = "black",
    linewidth = 1.2, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, RAIN_low),
    linetype = "dashed",
    color = "black",
    linewidth = 0.8, se = F
  ) +
  geom_smooth(
    data = wx_summarise,
    aes(dpp, RAIN_high),
    linetype = "dashed",
    color = "black",
    linewidth = 0.8, se = F
  ) +
  labs(
    x = "Days",
    y = "Precipitation (mm)"
  ) +
  theme_bw()+
  theme(text = element_text(size = 10),
        axis.title = element_text(face = "bold", size = 12))

tmean_plot+rain_plot+vpd_plot
```



```{r}
lat_plot = epi %>% 
  ggplot(aes(latitude))+
  geom_histogram(colour = "white", fill = "#002f61", bins = 12)+
   labs(
    x = "Latitude",
    y = "Frequency"
  ) +
  theme_bw()+
  theme(text = element_text(size = 10),
        axis.title = element_text(face = "bold", size = 12))



lon_plot = epi %>% 
  ggplot(aes(longitude))+
  geom_histogram(colour = "white", fill = "#002f61", bins = 12)+
   labs(
    x = "Longitude",
    y = "Frequency"
  ) +
  theme_bw()+
  theme(text = element_text(size = 10),
        axis.title = element_text(face = "bold", size = 12))
```

```{r}
weather_data <- epi %>%
  mutate(
    # Criando a data permitida diretamente com base no estado
    allowed_date = case_when(
      state == "MT" ~ as.Date(paste0(format(planting_date, "%Y"), "-09-16")), 
      state == "MS" ~ as.Date(paste0(format(planting_date, "%Y"), "-09-16")),
      state == "GO" ~ as.Date(paste0(format(planting_date, "%Y"), "-09-25")),
      state == "PR" ~ as.Date(paste0(format(planting_date, "%Y"), "-09-11")),
      state == "TO" ~ as.Date(paste0(format(planting_date, "%Y"), "-10-01")),
      state == "DF" ~ as.Date(paste0(format(planting_date, "%Y"), "-10-01")),
      state == "BA" ~ as.Date(paste0(format(planting_date, "%Y"), "-10-01")),
      state == "MG" ~ as.Date(paste0(format(planting_date, "%Y"), "-10-01")) 
    ),
    # Ajustando o ano da allowed_date para casos de janeiro e fevereiro
    allowed_date = if_else(
      format(planting_date, "%m") %in% c("01", "02"), 
      as.Date(paste0(as.numeric(format(planting_date, "%Y")) - 1, "-", format(allowed_date, "%m-%d"))),
      allowed_date
    ),
    # Calculando a diferença em dias
    days_difference = as.numeric(planting_date - allowed_date)
  )

days_plot = weather_data %>% 
ggplot(aes(days_difference)) +
  geom_histogram(color = "white", fill= "#002f61", bins = 12)+
  labs(
    x = "Days",
    y = "Frequency") +
  theme_bw()+
  theme(text = element_text(size = 10),
        axis.title = element_text(face = "bold", size = 12))
```

```{r}

(tmean_plot+rain_plot+vpd_plot)/(days_plot+lat_plot+lon_plot)

ggsave("fig/covariates.png", dpi = 600, bg = "white", height = 6, width = 10)
```

```{r, echo=TRUE, eval=FALSE}
library(dplyr)
library(lubridate)

# -------------------------------------------------------------------
# PARÂMETROS (iguais aos do DLNM)
# -------------------------------------------------------------------
END_DAP <- 85      # dia da avaliação final (R6)
LAG_MAX <- 70      # janela retrospectiva (dias antes do END_DAP)

# definição das fases (em dias ANTES do R6)
LATER_MAX_LAG   <- 40    # 0–40 dias antes do R6
EARLIER_MIN_LAG <- 41    # 41–70 dias antes do R6
EARLIER_MAX_LAG <- 70

# -------------------------------------------------------------------
# 1) ANO ÂNCORA DA EPIDEMIA
#    (usar o MESMO ano usado no DLNM / severidade)
# -------------------------------------------------------------------
epi_year_tbl <- epi_use %>%
  transmute(
    epi_id,
    epi_year = year          # ano da epidemia (plantio/estudo)
  )

# -------------------------------------------------------------------
# 2) JUNTAR ANO ÂNCORA AO CLIMA DIÁRIO
# -------------------------------------------------------------------
wx_use2 <- wx_use %>%
  left_join(epi_year_tbl, by = "epi_id")

# -------------------------------------------------------------------
# 3) SELECIONAR JANELA EFETIVA DO DLNM
#    (END_DAP - LAG_MAX) até END_DAP
# -------------------------------------------------------------------
start_dpp <- END_DAP - LAG_MAX   # ex.: 85 - 70 = 15

wx_win <- wx_use2 %>%
  filter(dpp >= start_dpp, dpp <= END_DAP) %>%
  mutate(
    lag_day = END_DAP - dpp,   # dias ANTES do R6
    phase = case_when(
      lag_day <= LATER_MAX_LAG ~ "Later",
      lag_day >= EARLIER_MIN_LAG & lag_day <= EARLIER_MAX_LAG ~ "Earlier"
    )
  )

wx_win <- wx_win %>%
  distinct(epi_id, epi_year, dpp, .keep_all = TRUE)

# (checagem opcional)
# wx_win %>% count(phase, lag_day) %>% arrange(phase, lag_day)

# -------------------------------------------------------------------
# 4) CALCULAR P90 GLOBAL (mesma referência para todos os anos)
# -------------------------------------------------------------------
p90_tbl <- wx_win %>%
  summarise(
    p90_tmean = quantile(tmean, 0.70, na.rm = TRUE),
    p90_rain  = quantile(rain,  0.70, na.rm = TRUE),
    p90_vpd   = quantile(vpd,   0.70, na.rm = TRUE)
  )

p90_tmean <- p90_tbl$p90_tmean
p90_rain  <- p90_tbl$p90_rain
p90_vpd   <- p90_tbl$p90_vpd

# -------------------------------------------------------------------
# 5) FREQUÊNCIA DE DIAS ≥ P90 (por epidemia × fase)
# -------------------------------------------------------------------
freq_epi <- wx_win %>%
  mutate(
    hit_tmean = as.integer(tmean >= p90_tmean),
    hit_rain  = as.integer(rain  >= p90_rain),
    hit_vpd   = as.integer(vpd   >= p90_vpd)
  ) %>%
  group_by(epi_id, epi_year, phase) %>%
  summarise(
   # n_days  = n(),
    n_tmean = sum(hit_tmean, na.rm = TRUE),
    n_rain  = sum(hit_rain,  na.rm = TRUE),
    n_vpd   = sum(hit_vpd,   na.rm = TRUE),
    .groups = "drop"
  )



freq_epi %>% 
ggplot(aes(y = n_vpd, x =  as.factor(epi_year))) +
  stat_halfeye(fill = "#ffc425", alpha = 0.7)+
  #geom_point()+
  ##geom_line()+
  facet_wrap(~phase)+
  coord_flip()

unique(freq_epi$epi_year)

freq_epi$epi_year = freq_epi$epi_year-1

freq_epi %>% 
  group_by(epi_year,phase) %>% 
  summarise(
    n_tmean = mean(n_tmean),
    n_rain = mean(n_rain),
    n_vpd = mean(n_vpd)
  )

max(freq_epi$n_vpd)
unique(wx_win$epi_id)
unique(wx_win$lag_day)

wx_win %>% 
  ggplot(aes(lag_day))+
  geom_histogram()

writexl::write_xlsx(wx_win,"data/wx_win.xlsx")


```
# Model simulation
```{r, echo=TRUE, eval=FALSE}
library(dplyr)
library(purrr)
library(tidyr)
library(dlnm)
library(glmmTMB)
library(epiR)

# ---------------- PARÂMETROS GERAIS ----------------
END_DAP   <- 85
LAG_MAX   <- 85
SEPARATOR <- LAG_MAX

LAG_BAND_A <- 0:40
LAG_BAND_B <- 41:LAG_MAX

# grades de teste
df_var_grid <- c(3, 4, 5)
df_lag_grid <- c(3, 4, 5)

VARS_ALL <- c("tmax", "rain", "vpd","rh","tmean","t2mdew","tmin","rain_cum")

# ---------------------------------------------------
# FUNÇÕES AUXILIARES
# ---------------------------------------------------

build_pooled_series <- function(wx_long, var, sep_n = SEPARATOR) {
  ids <- unique(wx_long$epi_id)
  unlist(lapply(ids, function(id) {
    v <- wx_long |>
      filter(epi_id == id) |>
      arrange(dpp) |>
      pull(.data[[var]])
    c(v, rep(NA_real_, sep_n))
  }))
}

extract_last_cb_row <- function(x, cb_template) {
  cb <- crossbasis(
    x, lag = LAG_MAX,
    argvar = attr(cb_template, "argvar"),
    arglag = attr(cb_template, "arglag")
  )
  as.numeric(cb[length(x), ])
}

build_design_for_var <- function(wx_long, var, cb_template, prefix) {
  X <- wx_long |>
    group_by(epi_id) |>
    summarise(cb = list(extract_last_cb_row(.data[[var]], cb_template)),
              .groups = "drop")
  p  <- length(X$cb[[1]])
  nm <- paste0(prefix, seq_len(p))
  X |>
    mutate(cb = lapply(cb, setNames, nm)) |>
    unnest_wider(cb)
}

# ---------------------------------------------------
# LOOP PRINCIPAL
# ---------------------------------------------------

results <- list()
model_id <- 1

for (DF_VAR in df_var_grid) {
  for (DF_LAG in df_lag_grid) {
    
    for (k in 1:length(VARS_ALL)) {
      var_sets <- combn(VARS_ALL, k, simplify = FALSE)
      
      for (VARS in var_sets) {
        
        # ---- construir cross-basis ----
        cb_templates <- list()
        for (v in VARS) {
          x_pool <- build_pooled_series(wx_use, v)
          cb_templates[[v]] <- crossbasis(
            x_pool,
            lag = LAG_MAX,
            argvar = list(fun = "ns", df = DF_VAR),
            arglag = list(fun = "ns", df = DF_LAG)
          )
        }
        
        # ---- design matrix ----
        X_list <- map(
          VARS,
          ~ build_design_for_var(
            wx_use,
            .x,
            cb_templates[[.x]],
            prefix = paste0("cb_", .x, "_")
          )
        )
        
        X <- reduce(X_list, left_join, by = "epi_id")
        
        dat <- epi_use |>
          select(epi_id, sev, siteyear) |>
          left_join(X, by = "epi_id") |>
          mutate(sev = sev / 100)
        
        cb_terms <- paste(names(dat)[grepl("^cb_", names(dat))], collapse = " + ")
        fml <- as.formula(paste0("sev ~ 1 + (1|siteyear) + ", cb_terms))
        
        # ---- ajuste do modelo ----
        fit <- try(
          glmmTMB(
            formula = fml,
            data = dat,
            family = beta_family(link = "logit")
          ),
          silent = TRUE
        )
        
        if (inherits(fit, "try-error")) next
        
        pred <- predict(fit, type = "response")
        ccc  <- epi.ccc(dat$sev, pred)
        
        results[[model_id]] <- tibble(
          model_id = model_id,
          DF_VAR   = DF_VAR,
          DF_LAG   = DF_LAG,
          vars     = paste(VARS, collapse = " + "),
          CCC      = ccc$rho.c[1],
          Cb       = ccc$C.b[1],
          rho      = ccc$rho[1]
        )
        
        model_id <- model_id + 1
      }
    }
  }
}

# ---------------------------------------------------
# RESULTADOS FINAIS
# ---------------------------------------------------

results_df <- bind_rows(results) |>
  arrange(desc(CCC))

top5_models <- results_df |> slice(1:20)

  top5_models$CCC
  
  results_df$CCC1 = results_df$CCC$est
  results_df$rho1 = results_df$rho$est
  
  
  writexl::write_xlsx(results_df,"data/results_df2.xlsx")
  
  results_df %>% 
    ggplot(aes(Cb))+
    geom_histogram()
  
  results_df %>% 
    filter(CCC >=0.80)

  library(stringr)  
  
results_df |>
  filter(str_detect(vars, "\\brain_cum\\b")) |>
  mutate(
    n_vars = str_count(vars, "\\+") + 1
  ) |>
  filter(n_vars <= 4)


results_df %>% 
  filter(model_id == 1647) %>% 
  summarize(
    CCC = CCC1
  )


results_df %>% 
  filter(CCC1 >= 0.8) %>% 
  filter(str_detect(vars, "\\brain_cum\\b")) |>
  mutate(
    n_vars = str_count(vars, "\\+") + 1
  ) |>
  filter(n_vars <= 4)
```



